---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

# Reproducible Research: Peer Assessment 1
**An analysis of personal activity data from an anonymous individual collected during 
October and November 2012.**

In this report we consider activity data collected from an anonymous individual. The
data were collected at five-minute intervals during the day, although there are a number
of days missing data. We report some summaries of the individual's activity, demonstrate
the negative impact of imputing the missing values, and finally contrast the activity
profiles of weekdays and weekends.

## Loading and preprocessing the data
```{r loaddata}
raw.data <- read.csv("activity.csv", colClasses=c("numeric", "character", "numeric"))
raw.data$date <- as.Date(raw.data$date)
 ```
Having read in the raw data and converted date to Date class, it has the structure:
```{r structureData}
str(raw.data)
```



## What is mean total number of steps taken per day?
First, let's aggregate the data for each day:
```{r totaldaily}
total.daily.steps <- aggregate(steps ~ date, raw.data, sum)
```
This gives data suitable for answering this question:
```{r showtotaldaily}
head(total.daily.steps)
```
The figure below shows the histogram of total steps taken each day.
```{r histdaily}
h <- hist(total.daily.steps$steps, breaks=15, xlab="Total daily number of steps",
        col="lightgrey", main=NA)
mean.value      <- mean(total.daily.steps$steps)
median.value    <- median(total.daily.steps$steps)
```

The **mean** value of the total number of steps taken each day is `r sprintf("%.1f", mean.value)` and
the **median** value is `r sprintf("%.1f", median.value)`.



## What is the average daily activity pattern?
Now we want to aggregate the data over the days to see the average activity profile
during the day:
```{r averagedaily}
average.daily.steps <- aggregate(steps ~ interval, raw.data, mean)
with(average.daily.steps, plot(interval, steps, type="l", ylab="Average number of steps",
                                main="Average daily activity pattern"))
max.interval <- with(average.daily.steps, interval[which.max(steps)])
abline(v=max.interval, lty=2)
legend("topright", sprintf("interval %d", max.interval), lty=2)
```

Interval `r max.interval`, on average, contains the maximum number of steps.


## Imputing missing values
```{r findmissing}
cc                  <- complete.cases(raw.data)
number.incomplete   <- length(cc)-sum(cc)
```

There are `r number.incomplete` missing values in the data, which corresponds to
`r sprintf("%.1f%%", 100*number.incomplete/length(cc))` of cases.
Checking how these are distributed within the data,
```{r showmissingdates}
with(raw.data, table(date[is.na(steps)]))
```
we see that the missing values are concentrated in 8 days for which all values are absent
(288 being the number of five-minute intervals in a day).
We will use the mean value (for an interval) to replace missing (steps) data.
We know it is entire days that are missing,
rather than random intervals here and there, but we will not require this for the imputation of the missing
values. Neither will we assume the ordering of the intervals for each day.

The basic strategy here is:  

1. Find the locations (and hence the interval values) of the missing values.  
2. Find the indices of the relevant intervals in the daily average data.  
3. Use these indices to extract the (replacement) values from the daily average data and
substitute back into the data in the locations of the associated NA values.  

A completely missing day would thus be replaced exactly by a copy of the average daily activity profile.

```{r imputing}
steps.na    <- is.na(raw.data$steps)
#now for each interval with NA steps, find the index of that interval in the average:
ind.in.avg  <- sapply(raw.data$interval[steps.na], function(x) which(x == average.daily.steps$interval))
#the logical vector of NA values now corresponds to indices of values in average.daily.steps
raw.data.no.missing                 <- raw.data
raw.data.no.missing$steps[steps.na] <- average.daily.steps$steps[ind.in.avg]
number.missing.in.duplicated        <- sum(is.na(raw.data.no.missing$steps)) #check for any still missing 
```
There are `r number.missing.in.duplicated` missing values in the duplicate data.

```{r histdaily2}
total.daily.steps2 <- aggregate(steps ~ date, raw.data.no.missing, sum)
h2 <- hist(total.daily.steps2$steps, breaks=15, xlab="Total daily number of steps\n(Missing values substituted)",
        col="lightgrey", main=NA)
mean.value2      <- mean(total.daily.steps2$steps)
median.value2    <- median(total.daily.steps2$steps)
```
The **mean** value of the total number of steps taken each day, after imputing missing values,
is `r sprintf("%.1f", mean.value2)` and the **median** value is `r sprintf("%.1f", median.value2)`.
By imputing missing values in this manner, we have biased the distribution towards the mean. The mean is unchanged
from the previous calculation (with missing values) but the median has now converged to the mean. The effect is
clear when comparing the two histograms, with all the imputed observations falling into a single bin. This is
confirmed by printing the counts for each histogram:
```{r histcounts}
print(h$counts)
print(h2$counts)
```
All 8 missing days have been added into a single bin, significantly distorting the distribution.

## Are there differences in activity patterns between weekdays and weekends?
Here we will look to see if there is a difference in activity throughout the day between weekdays
and weekends.

```{r weekdays}
daynames <- weekdays(raw.data.no.missing$date)
#create new factor in data.frame
raw.data.no.missing$weekend <- factor(ifelse( (daynames=="Saturday" | daynames=="Sunday"), 
                                                "weekend", "weekday" ))
rm("daynames")

#and now calculate the average number of steps throughout a day, but conditioned on weekend/weekday
average.daily.no.missing <- aggregate(steps ~ interval + weekend, raw.data.no.missing, mean)

library(lattice)
xyplot(steps ~ interval | weekend, average.daily.no.missing, 
        type="l", 
        layout=c(1, 2), 
        xlab="Interval", 
        ylab="Number of steps")
```

It is clear from the plot that weekdays have a sharper early burst of activity. Also, it appears to start 
slightly earlier than the weekends. The plots also show the subject is more active throughout the day at weekends
and the activity lasts longer into the evening than on a weekday.
